---
title: "Lecture_Assign_DataExplore_Deterministic"
author: "Johanna Ravenhurst"
date: "2022-09-28"
output: html_notebook
---

Data Exploration and Deterministic Functions
https://michaelfrancenelson.github.io/environmental_data/assignments/individual_03_data_exploration_deterministic_functions.html

```{r}
#Installing packages
#install.packages("here")
require(here)

```
#here package:
#Use the here package instead of setwd() to set the working directory when working within an Rproject

```{r}
here()

#use here to point to a subdirectory within Rproject - subdirectory is "data"
#you can add another subdirectory after "data" if needed
dat_habitat <- read.csv(here("data", "hab.sta.csv"))

head(dat_habitat)
names(dat_habitat)

#If needed: reality check to see if file exists in directory
file.exists(here("data", "hab.sta.csv"))
```
#HISTOGRAMS

#Plot histograms of the following terrain variables:
elevation (m) numeric, 85-872
aspect (degrees) numeric 0-360
slope (percent slope %) 0-110

```{r}

par(mfrow = c(1, 3))
elev_hist <-
  hist (dat_habitat$elev,
      main = "Histogram of Habitat Elevation\nBy Johanna Ravenhurst",
      col =
        adjustcolor(col = "steelblue", alpha.f =.4),
      border = "blue",
      xlab = "Elevation")

slope_hist <-
hist (dat_habitat$slope,
      main = "Histogram of Habitat Slope\nBy Johanna Ravenhurst",
      col =
        adjustcolor(col = "steelblue", alpha.f =.4),
      border = "blue",
      xlab = "Slope")

aspect_hist <-
hist (dat_habitat$aspect,
      main = "Histogram of Habitat Aspect\nBy Johanna Ravenhurst",
      col =
        adjustcolor(col = "steelblue", alpha.f =.4),
      border = "blue",
      xlab = "Aspect",
      breaks=15)

#NOTE: Aspect ranges from 0-360 and the default bins went up to 400, increasing the number of breaks fixed this issue to provide a more accurate visualization of the data

png(filename = here("histograms_habitat_terrain.png"), width = 800, height = 900)
par(mfrow = c(2, 2))

hist (dat_habitat$elev,
      main = "Histogram of Habitat Elevation\nBy Johanna Ravenhurst",
      col =
        adjustcolor(col = "steelblue", alpha.f =.4),
      border = "blue",
      xlab = "Elevation")

hist (dat_habitat$slope,
      main = "Histogram of Habitat Slope\nBy Johanna Ravenhurst",
      col =
        adjustcolor(col = "steelblue", alpha.f =.4),
      border = "blue",
      xlab = "Slope")

hist (dat_habitat$aspect,
      main = "Histogram of Habitat Aspect\nBy Johanna Ravenhurst",
      col =
        adjustcolor(col = "steelblue", alpha.f =.4),
      border = "blue",
      xlab = "Aspect", 
      breaks=15)
dev.off()

elev_hist
slope_hist
aspect_hist

#Not sure how to find the number of observations for a given variable
summary(dat_habitat$slope)
table(dat_habitat$slope)
dat_habitat
```
#Scatterplots
#Create scatterplots of total basal area and the terrain variables (consult the metadata file to see which column(s) you need). Basal area should be on the y-axis.

```{r}

elev_plot <-
plot(x = dat_habitat$elev, y = dat_habitat$ba.tot, main = "Basal Area and Elevation", xlab = "Elevation", ylab = "Total Basal Area",
     cex = 0.5)

slope_plot <-
plot(x = dat_habitat$slope, y = dat_habitat$ba.tot, main = "Basal Area and Slope", xlab = "Slope", ylab = "Total Basal Area",
     cex = 0.5)

aspect_plot <-
plot(x = dat_habitat$aspect, y = dat_habitat$ba.tot, main = "Basal Area and Aspect", xlab = "Aspect", ylab = "Total Basal Area",
     cex = 0.5)

#There are many points, some of which partially overlap, in the scatterplots. You may want to experiment with the size of individual points to see if you can find an optimal size that allows you to see all of them. Try different values of the cex argument in your plot() call.


```

#Fitting Linear Functions

```{r}

# Calculates the value of y for a linear function, given the coordinates
# of a known point (x1, y1) and the slope of the line.
line_point_slope = function(x, x1, y1, slope)
{
  get_y_intercept = 
    function(x1, y1, slope) 
      return(-(x1 * slope) + y1)
  
  linear = 
    function(x, yint, slope) 
      return(yint + x * slope)
  
  return(linear(x, get_y_intercept(x1, y1, slope), slope))
}

#Hint: Choose a different color for your lines. Experiment with the col argument to plot().

plot( x = dat_habitat$elev, y = dat_habitat$ba.tot, main = "Visually-estimated linear model fit\nBasal Area and Elevation", xlab = "Elevation", ylab = "Total Basal Area",
      cex = 0.5)

curve(line_point_slope(x, x1 = 400, y1 = 40, slope = -0.07), add = TRUE, col = "dark red")


plot(
  x = dat_habitat$slope, 
  y = dat_habitat$ba.tot, 
  main = "Visually-estimated linear model fit\nBasal Area and Slope", 
  xlab = "Slope", 
  ylab = "Total Basal Area",
  cex = 0.5)

curve(line_point_slope(x, x1 = 50, y1 = 25, slope = 0.05), add = TRUE, col = "dark red")


plot(
  x = dat_habitat$aspect, 
  y = dat_habitat$ba.tot, 
  main = "Visually-estimated linear model fit\nBasal Area and Aspect", 
  xlab = "Aspect", 
  ylab = "Total Basal Area",
  cex = 0.5)


curve(line_point_slope(x, x1 = 200, y1 = 25, slope = 0.0), add = TRUE, col = "dark red")


```
CREATE FIGURE WITH ALL HISTOGRAMS AND SCATTERPLOTS

```{r}
png(filename = here("hist_scatter_habitat_terrain.png"), width = 1000, height = 900)
par(mfrow = c(2, 3))

hist (dat_habitat$elev,
      main = "Histogram of Habitat Elevation\nBy Johanna Ravenhurst",
      col =
        adjustcolor(col = "steelblue", alpha.f =.4),
      border = "blue",
      xlab = "Elevation")

hist (dat_habitat$slope,
      main = "Histogram of Habitat Slope\nBy Johanna Ravenhurst",
      col =
        adjustcolor(col = "steelblue", alpha.f =.4),
      border = "blue",
      xlab = "Slope")

hist (dat_habitat$aspect,
      main = "Histogram of Habitat Aspect\nBy Johanna Ravenhurst",
      col =
        adjustcolor(col = "steelblue", alpha.f =.4),
      border = "blue",
      xlab = "Aspect", 
      breaks=15)

plot( x = dat_habitat$elev, y = dat_habitat$ba.tot, main = "Visually-estimated linear model fit\nBasal Area and Elevation", xlab = "Elevation", ylab = "Total Basal Area",
      cex = 0.5)

curve(line_point_slope(x, x1 = 400, y1 = 40, slope = -0.07), add = TRUE, col = "dark red")


plot(
  x = dat_habitat$slope, 
  y = dat_habitat$ba.tot, 
  main = "Visually-estimated linear model fit\nBasal Area and Slope", 
  xlab = "Slope", 
  ylab = "Total Basal Area",
  cex = 0.5)

curve(line_point_slope(x, x1 = 50, y1 = 25, slope = 0.05), add = TRUE, col = "dark red")


plot(
  x = dat_habitat$aspect, 
  y = dat_habitat$ba.tot, 
  main = "Visually-estimated linear model fit\nBasal Area and Aspect", 
  xlab = "Aspect", 
  ylab = "Total Basal Area",
  cex = 0.5)


curve(line_point_slope(x, x1 = 200, y1 = 25, slope = 0.0), add = TRUE, col = "dark red")


dev.off()
```
