---
title: "Lecture DataCamp Assignment"
author: "Johanna Ravenhurst"
date: "2022-09-12"
output: html_notebook
---

#Create:

#    A variable a that contains the text of your first name.
#    A variable b1 that contains the number 45.6
#    A variable b2 that contains the text “45.6”
#    A variable c1 that contains the sequence of integers from 0 to 3

```{r}
#Code for questions 1-6
a <- "Johanna"
b1 <- 45.6
b2 <- "45.6"
c1 <- 0:3
a
b1
b2
c1
class(a)
class(b1)
class(b2)
class(c1)
b1+b2
b1+c1

```

```{r}
#Code for questions 7-9
#Create a vector called v1 that contains a sequence of integers from -2 to 2.
v1 <- c(-2:2)
v1

#Now, use v1 to create a new vector called v2 whose elements are the elements of v1 multiplied by 3. 
v2 <- c(v1*3)
v2

#Finally, calculate the sum of all the elements in v2.
sum(v2)


```

```{r}
#Code for questions 10-11
#Create a vector called vec_4 whose elements are the integers from 1 to 12
vec_4 <- c(1:12)
vec_4

#Create a matrix mat_1 from vec_4 that has three rows and four columns. The values in mat_1 should be sequentially increasing by row.
mat_1 <- matrix(vec_4, byrow= TRUE, nrow = 3)
mat_1

#Create a matrix mat_2 from vec_4 that has three rows and four columns. The values in mat_2 should be sequentially increasing by column
mat_2 <- matrix(vec_4, byrow= FALSE, nrow = 3 )
mat_2

```
Lists

Create a list, named my_list_1 with following three elements:

    first element is numeric: 5.2
    second element is a string: “five point two”
    third element is a vector of all integers from 0 to 5 [how do you do this?]

Name the elements in my_list_1:

    “two”
    “one”
    “three”

```{r}
#Code for questions 12-14

my_list_1 <- list (two = 5.2, one = "five point two", three = 0:5)

OR
my_list_3 <- list (5.2, "five point two", 0:5)
names(my_list_3) <- c("two", "one","three")

#select the third element of the list
my_list_1[[3]]

#select the list element with the name "one"
my_list_1$one
my_list_1[["one"]]


```
#Logical Tests and Subsetting

```{r}
#Code for questions 15-16
my_vec = rep(1:3, 5)
my_vec

#Use the logical equality test operator == to create a vector, my_bool_vec, of Boolean values from my_vec.

   # my_bool_vec should be the same length as my_vec.
   # my_bool_vec should have TRUE values in the positions where my_vec has values of 3.

my_bool_vec <- my_vec == 3
my_bool_vec


#You can run the following code to check that you have the correct values in my_bool_vec:
data.frame(my_vec, my_bool_vec)

#Use my_bool_vec to retrieve all of the elements of my_vec that have a value of 3.

my_vec[my_bool_vec == TRUE]

```

