---
title: "InClass_DataFilePractice"
author: "Johanna Ravenhurst"
date: "2022-09-20"
output: html_document
---

Use here() and read.csv() to read the three files into data.frame objects called:

    dat_catrate
    dat_delomys
    dat_rope

Use head() to view the first few lines of each data frame.

Check how many columns contain numeric data in each of the 3 data frames.



```{r}
library(here)

dat_catrate <- read.csv(here("data", "catrate.csv"))
dat_delomys <- read.csv(here("data", "delomys.csv"))
dat_rope <- read.csv(here("data", "rope.csv"))

head(dat_catrate)
head(dat_delomys)
head(dat_rope)


```

You can plot histograms for any column that contains numeric data. For data sets with 2 or more numeric columns you can create scatterplots.

```{r}

plot(x = dat_catrate$pond, y = dat_catrate$success)

plot(x = dat_rope$p.cut, y = dat_rope$p.strength, main="Johanna Ravenhurst's Plot", xlab= "Rope p.cut", ylab = "Rope strength")

png(filename = here("inclass_scatter_rope.png"), width = 800, height = 600)
plot(x = dat_rope$p.cut, y = dat_rope$p.strength, main="Johanna Ravenhurst's Plot")
dev.off()
```


```{r}
#After class - attempt at making a histogram 
# use /n to start new line in title
#use col to adjust the color
#adjust color makes it lighter if less than 1

hist (dat_delomys$body_length,
      main = "Histogram of Body Length\nBy Johanna Ravenhurst",
      col =
        adjustcolor(col = "steelblue", alpha.f =.4),
      border = "blue",
      xlab = "Body Length")

#Since these data are skewed, I am going to try limited the x values with xlim and adjusting the number of bins

hist (dat_delomys$body_length,
      main = "Histogram of Body Length\nBy Johanna Ravenhurst",
      col =
        adjustcolor(col = "steelblue", alpha.f =.4),
      border = "blue",
      xlab = "Body Length",
      xlim = c(0, 200),
      breaks = 10)

boxplot(body_length ~ sex, data = dat_delomys)

par(mfrow = c(1, 2))
hist (dat_delomys$body_length,
      main = "Histogram of Body Length\nBy Johanna Ravenhurst",
      col =
        adjustcolor(col = "steelblue", alpha.f =.4),
      border = "blue",
      xlab = "Body Length",
      xlim = c(0, 300),
      breaks = 20)
boxplot(body_length ~ sex, data = dat_delomys, 
        main= "Boxplot of Body Length\nBy Johanna Ravenhurst", 
        col= "red", alpha.f = .5)

```


Histogram examples - from In-Class R Demos and Examples
https://michaelfrancenelson.github.io/environmental_data/#In-Class_R_Demos_and_Examples

```{r}
#FANCY HISTOGRAMS
#You can use the special character \n to insert a line break into a title or axis label.
#You can use the col argument to specify a color for the bars.
#You can use the border argument to specify the color of the outlines of the bars.
#You can use the adjustcolor() function to make a color lighter by specifying an alpha value of less than 1.


require(palmerpenguins)

hist(
  penguins$bill_length_mm,
  main = "Hist 'o Gram of Bill Length\nBy Mike Nelson",
  col = 
    adjustcolor(col = "steelblue", alpha.f = .4),
  border = "red",
  xlab = "Bill Length (in mm)")

```

```{r}
#If a histogram has some very short bins at high values of x, you can truncate the display using the xlim argument. For example, if you had some data stored in a data frame and you wanted to make a histogram of the column called wingspan:

#This doesn’t look great:

hist(
  dat$wingspan,
  main = "Histogram of Wingspan",
  xlab = "wingspan (cm)")

#I can truncate the x-values to be between 0 and 10 using xlim.

#I’ll also make more bins using the breaks argument.

  #Note that R considers the breaks argument to be a suggestion… Telling it to create 30 bins won’t always result in 30 bins. You can experiment with different numbers of bins until you find a number that works for your plot.

hist(
  dat$wingspan,
  main = "Histogram of Wingspan",
  xlab = "wingspan (cm)",
  xlim = c(0, 10),
  breaks = 30)
```

