---
title: "Lab 1 R Fundamentals 1"
author: "Johanna Ravenhurst"
date: "2022-09-14"
output: html_notebook
---



```{r}
#Expressions and Strings - Q1

c(1, 2, 3)
"c(1, 2, 3)"

```

```{r}
#Variables - Q2-4

c_1 = c(1, 2, 3)
c_2 = "c(1, 2, 3)"
c_1
c_2



```
```{r}
#Matrices 1 - Q5-6

#Q6 (2 pts.): Write R code to retrieve the element of mat_1 that has a value of 3.

my_vec <- c(1:6)
my_vec
mat_1 = matrix(my_vec, nrow = 3)

#Q5 (1 pt.): What are the dimensions of the matrix (i.e. how many rows and columns)?
# 3 rows, 2 columns

mat_1

#Q6 (2 pts.): Write R code to retrieve the element of mat_1 that has a value of 3.

mat_1[3,1]


```

```{r}
#Matrices 2 - Q7-11

#You will use my_vec from the previous question again.
my_vec


#Create a matrix mat_2 that has two rows and three columns using my_vec. Do not use the c() or rep() functions.
mat_2 = matrix(my_vec, nrow = 2, ncol = 3)
  

#Create a matrix mat_3 that has three rows and two columns using my_vec. Do not use the c() or rep() functions.
mat_3 = matrix(my_vec, nrow = 3, ncol = 2)

mat_1
mat_2
mat_3

#Q10 (1 pt.): Using my_vec, create a matrix, mat_4. mat_4 must have a total number of elements that is not a multiple of 3.

mat_4 = matrix(my_vec, nrow=2, ncol=1)
mat_4

mat_5 = matrix(my_vec, nrow=10, ncol=5)
mat_5





```

```{r}
#List subsetting - Q12-14

#Create a list, named my_list_1 with following three elements:

    #first element is numeric: 5.2
    #second element is a string “five point two”
    #third element is a vector of all integers from 0 to 5.

my_list_1 <- list (two = 5.2, one = "five point two", three = 0:5)

#Name the elements in my_list_1:

   # “two”
   # “one”
   # “three”



#Run the following 8 lines of code.

my_list_1[[1]]
my_list_1[[as.numeric("1")]]
my_list_1[["1"]]
my_list_1[["one"]]
my_list_1$one
my_list_1$"one"
my_list_1$1
my_list_1$"1"



  #Q12 (8 pts.): For each of the 8 lines, answer the following: A. Did the line return a 1: value, 2: error, or 3: NULL? B. What type of subsetting operation was used (or attempted)? C. If it did not return an error describe, in ordinary English, a plausible explanation of how R could have performed the subsetting.

   # Q13 (2 pts.): Identify which lines produced the string output "five point two" and explain why.

   # Q14 (1 pt.): Identify which lines produced NULL output and explain why.


```



#Vectors (one dimensional array): can hold numeric, character or logical values. The elements in a vector all have the same data type.

#Matrices (two dimensional array): can hold numeric, character or logical values. The elements in a matrix all have the same data type.

#Data frames (two-dimensional objects): can hold numeric, character or logical values. Within a column all elements have the same data type, but different columns can be of different data type.

#A list in R allows you to gather a variety of objects under one name (that is, the name of the list) in an ordered way. These objects can be matrices, vectors, data frames, even other lists, etc. It is not even required that these objects are related to each other in any way.
